---
title: "Neutrophil Function Demographics Analysis"
author: "Elise Hickman"
date: "2023-11-14"
output: html_document
---

# Prepare workspace

```{r message = FALSE}
# Clear global environment
rm(list = ls(all.names = TRUE))

# Load packages
library(openxlsx) # for data import 
library(table1) # to make tables
library(tidyverse) # for data cleaning and organization
library(janitor) # for data cleaning
library(ggplot2) # for bar plot
library(ggsignif) # for bar plot
library(corrplot) # for correlation plot
library (RColorBrewer) # for correlation plot

# Set graphing theme
theme_set(theme_bw())
```

# Import and clean data

Seahorse data:
```{r}
seahorse_data <- data.frame(read.xlsx("1_InputData/2021_12_14 Seahorse Demographics Analysis.xlsx"))

# Make sure variables are numeric that should be numeric. 
# They sometimes read in as characters because of how the study coordinators extract them from the database.
seahorse_data$Age <- as.numeric(seahorse_data$Age)
seahorse_data$BMI <- as.numeric(seahorse_data$BMI)
```

Phagocytosis data:

```{r}
phagocytosis_data <- data.frame(read.xlsx("1_InputData/2021_12_14 Phagocytosis Demographics Analysis.xlsx"))

# Make sure variables are numeric that should be numeric. 
# They sometimes read in as characters because of how the study coordinators extract them from the database.
phagocytosis_data$Age <- as.numeric(phagocytosis_data$Age)
phagocytosis_data$BMI <- as.numeric(phagocytosis_data$BMI)
```

# Make demographics tables

Seahorse data:
```{r}
# Change labels
seahorse_data$Race <- factor(seahorse_data$Race, 
                          levels=c("W", "B", "A", "MO"), 
                          labels=c("White", "Black", "Asian", "Mixed/Other"))

seahorse_data$Sex <- factor(seahorse_data$Sex, 
                             levels=c("F", "M"),
                             labels=c("Female", "Male"))

# Make table
table1(~ Sex + Race + Hispanic + Age + BMI,
       data = seahorse_data, 
       render.missing = NULL)
```

Phagocytosis data:
```{r}
# Change labels
phagocytosis_data$Race <- factor(phagocytosis_data$Race, 
                             levels=c("W", "B", "A", "MO"), 
                             labels=c("White", "Black", "Asian", "Mixed/Other"))

phagocytosis_data$Sex <- factor(phagocytosis_data$Sex, 
                            levels=c("F", "M"),
                            labels=c("Female", "Male"))

# Make table
table1(~ Sex + Race + Hispanic + Age + BMI,
       data = phagocytosis_data, 
       render.missing = NULL)
```

# Assess normality of data

Seahorse data:
```{r}
# Statistical test of normality
seahorse_normality <- apply(seahorse_data[9:16], 2, shapiro.test)

# Create data frame to summarize results
seahorse_normality <- do.call(rbind.data.frame, seahorse_normality )
seahorse_normality <- format(seahorse_normality, scientific = FALSE)

# Add column to adjust for multiple hypothesis testing
seahorse_normality$p.value.adj <- p.adjust(seahorse_normality$p.value, "BH")

# Add column for normality conclusion
seahorse_normality <- seahorse_normality %>% mutate(normal = ifelse(p.value.adj < 0.05, F, T)) 

# Make vector of names for Seahorse metrics that are normally distributed.
normal_seahorse_metrics <- filter(seahorse_normality, normal == 'TRUE')
normal_seahorse_metrics = rownames(normal_seahorse_metrics)
```

Phagocytosis data:
```{r}
phagocytosis_normality <- shapiro.test(phagocytosis_data$Phagocytosis)
phagocytosis_normality$p.value # p = 0.1772016 means normally distributed
```

# Summary stats for Table 1

Seahorse:
```{r warning = FALSE}
# Define functions for summary
summary_functs <- lst(mean, sd)

# Store plus/minus character
x<-"\u00b1"
Encoding(x)<-"UTF-8"
plus_minus <- paste0(" ", x, " ", sep = "")

# All subjects together
seahorse_summarystats <- map_dfr(summary_functs, ~ summarize(seahorse_data, across(9:ncol(seahorse_data), .x, na.rm = TRUE)), .id = "statistic") %>%
  t() %>% as.data.frame() %>%
  row_to_names(1) %>%
  mutate(mean = round(as.numeric(mean), digits = 0)) %>%
  mutate(sd = round(as.numeric(sd), digits = 0)) %>%
  mutate(mean_sd = paste0(mean, plus_minus, sd, sep ="")) %>%
  select(mean_sd) %>%
  rownames_to_column("Metric") %>%
  rename("Total" = "mean_sd")

# Males
seahorse_data_male <- seahorse_data %>% filter(Sex == "Male")

seahorse_summarystats_male <- map_dfr(summary_functs, ~ summarize(seahorse_data_male, across(9:ncol(seahorse_data_male), .x, na.rm = TRUE)), .id = "statistic") %>%
  t() %>% as.data.frame() %>%
  row_to_names(1) %>%
  mutate(mean = round(as.numeric(mean), digits = 0)) %>%
  mutate(sd = round(as.numeric(sd), digits = 0)) %>%
  mutate(mean_sd = paste0(mean, plus_minus, sd, sep ="")) %>%
  select(mean_sd) %>%
  rownames_to_column("Metric") %>%
  rename("Male" = "mean_sd")


# Females
seahorse_data_female <- seahorse_data %>% filter(Sex == "Female")

seahorse_summarystats_female <- map_dfr(summary_functs, ~ summarize(seahorse_data_female, across(9:ncol(seahorse_data_female), .x, na.rm = TRUE)), .id = "statistic") %>%
  t() %>% as.data.frame() %>%
  row_to_names(1) %>%
  mutate(mean = round(as.numeric(mean), digits = 0)) %>%
  mutate(sd = round(as.numeric(sd), digits = 0)) %>%
  mutate(mean_sd = paste0(mean, plus_minus, sd, sep ="")) %>%
  select(mean_sd) %>%
  rownames_to_column("Metric") %>%
  rename("Female" = "mean_sd")

# Merge together and write out
summary_stats_final <- seahorse_summarystats %>%
  left_join(seahorse_summarystats_male, by = "Metric") %>%
  left_join(seahorse_summarystats_female, by = "Metric") 

write.xlsx(summary_stats_final, "2_OutputTables/Seahorse_SummStats.xlsx")
```

Phagocytosis:
```{r}
# All together
mean(phagocytosis_data$Phagocytosis)
sd(phagocytosis_data$Phagocytosis)

# Male
p_data_male <- phagocytosis_data %>% filter(Sex == "Male")
mean(p_data_male$Phagocytosis)
sd(p_data_male$Phagocytosis)

# Female
p_data_female <- phagocytosis_data %>% filter(Sex == "Female")
mean(p_data_female$Phagocytosis)
sd(p_data_female$Phagocytosis)
```

# Statistical tests for associations with demographic variables

## Sex

Seahorse:
```{r warning = FALSE}
# T-test for normally distributed data and Wilcox test for non-normally distributed data

# Create results data frame
seahorse_sex = data.frame()

# For loop that runs a t-test if the data are normally distributed and a wilcox test if the data are non-normally distributed.
for (i in 9:ncol(seahorse_data)) {
  if (names(seahorse_data[i]) %in% normal_seahorse_metrics == TRUE) {
    test <- t.test(as.formula(paste0(names(seahorse_data)[i], "~", "Sex", sep = "")), seahorse_data)
  } else {test <- wilcox.test(as.formula(paste0(names(seahorse_data)[i], "~", "Sex", sep = "")), seahorse_data) 
  }
  
  pval <- data.frame(test$p.value, row.names = noquote(paste0(names(seahorse_data[i]))))
  pval$method <- test$method
  
  seahorse_sex <- bind_rows(seahorse_sex, pval)
  
}

# View results
seahorse_sex

# Save results
seahorse_sex$metric <- rownames(seahorse_sex)
write.xlsx(seahorse_sex, "2_OutputTables/Seahorse_Sex_Results.xlsx")
```
Phagocytosis:
```{r}
phagocytosis_sex <- t.test(Phagocytosis ~ Sex, phagocytosis_data)
phagocytosis_sex$p.value
```

## Age

Spearman's correlation chosen because it can detect both linear and monotonic correlations, whereas Pearson's correlation only detects linear correlations. 

Seahorse:

```{r warning = FALSE}
seahorse_spearman_age <- apply(seahorse_data[, 9:16], 2, cor.test, seahorse_data$Age, method = "spearman")
seahorse_spearman_age <- as.data.frame(do.call(rbind, seahorse_spearman_age))
seahorse_spearman_age$metric <- rownames(seahorse_spearman_age)
seahorse_spearman_age <- apply(seahorse_spearman_age,2,as.character)

write.xlsx(seahorse_spearman_age, "2_OutputTables/Seahorse_Age_Results.xlsx")
```
Phagocytosis:
```{r warning = FALSE}
phagocytosis_spearman_age <- cor.test(phagocytosis_data$Age, phagocytosis_data$Phagocytosis, method = "spearman")
phagocytosis_spearman_age
```

## BMI

Seahorse:
```{r warning = FALSE}
seahorse_spearman_bmi <- apply(seahorse_data[, 9:16], 2, cor.test, seahorse_data$BMI, method = "spearman")
seahorse_spearman_bmi <- as.data.frame(do.call(rbind, seahorse_spearman_bmi))
seahorse_spearman_bmi$metric <- rownames(seahorse_spearman_bmi)
seahorse_spearman_bmi <- apply(seahorse_spearman_bmi,2,as.character)

write.xlsx(seahorse_spearman_bmi, "2_OutputTables/Seahorse_BMI_Results.xlsx")
```

Phagocytosis:
```{r}
phagocytosis_spearman_BMI <- cor.test(phagocytosis_data$BMI, phagocytosis_data$Phagocytosis, method = "spearman")
phagocytosis_spearman_BMI
```


# Plot sex differences in phagocytosis
```{r}
# To keep jitter consistent
set.seed(8016)

# Make plot
phagocytosis_plot <- ggplot(phagocytosis_data, aes(x = Sex, y = Phagocytosis, fill = Sex)) + 
  scale_fill_manual(values = c("tomato", "steelblue1")) +
  geom_boxplot(color = "black") +
  geom_jitter(position = position_jitter(0.15),
              shape = 19,
              size = 3) +
  geom_signif(comparisons = list(c("Female", "Male")),
              map_signif_level=TRUE,
              size = 1,
              textsize = 16,
              margin_top = 0.2) +
  ylim(0, 80000) +
  ylab("Phagocytosis (MFI)\n") +
  theme(legend.position = "None",
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 20, face = "bold", color = "black"),
        axis.title.y = element_text(size = 20, face = "bold", vjust = 0.5),
        axis.text.y = element_text(size = 18, color = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1))

# Write out
pdf("3_OutputFigures/PhagocytosisSexGraph.pdf",
    colormodel = "cmyk",
    width = 5,
    height = 5.5)

phagocytosis_plot

invisible(dev.off())

phagocytosis_plot
```

# Correlation between phagocytosis and seahorse metrics

## All subjects together 

```{r}
# Make phagocytosis data frame for merging
phagocytosis_data_trimmed <- subset(phagocytosis_data, TRUE, c("SubjectNumber", "Phagocytosis"))

# Combined Data
combined_data <- transform(merge(phagocytosis_data_trimmed, seahorse_data, by = "SubjectNumber"))

# Make data frame for just correlations
combined_data_trimmed <- combined_data[-c(3:9)]
combined_data_trimmed <- data.frame(combined_data_trimmed[, -1], row.names = combined_data_trimmed$SubjectNumber)
```

Make demographics table:

```{r}
table1(~ Sex + Race + Hispanic + Age + BMI,
       data = combined_data, 
       render.missing = NULL)
```

Make correlation plot:

```{r}
# Convert to data matrix
combined_data_trimmed_dm <- as.matrix(combined_data_trimmed)

# Create correlation data
correlations <- cor(combined_data_trimmed_dm, method = "spearman")

# To add significance to correlation plot, run this function
cor.mtest <- function(mat, ...) {
  mat <- as.matrix(mat)
  n <- ncol(mat)
  p.mat<- matrix(NA, n, n)
  diag(p.mat) <- 0
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      tmp <- cor.test(mat[, i], mat[, j], method = "spearman", exact = FALSE, ...)
      p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
    }
  }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}

p.mat_correlations <- cor.mtest(combined_data_trimmed_dm)

# Open graphical device
pdf("3_OutputFigures/NeutrophilFunctionCorrelations.pdf",
    colormodel = "cmyk")

# Make plot
corrplot(correlations, 
         method = "circle", 
         type = "upper",
         order = "original",
         tl.col = "black",
         p.mat = p.mat_correlations,
         sig.level = 0.05,
         tl.srt = 45,
         insig = "label_sig",
         pch.col = 'white',
         diag = FALSE,
         col=brewer.pal(n=10, name="RdYlBu"))

# Close Graphical Device
invisible(dev.off())
```

## Males

```{r}
# Filter to only males
combined_data_male <- filter(combined_data, Sex == "Male")

# Make data frame with only variables of interest for correlating
combined_data_male_trimmed <- combined_data_male[-c(3:9)]
combined_data_male_trimmed <- data.frame(combined_data_male_trimmed[, -1], row.names = combined_data_male_trimmed$SubjectNumber)

# Convert to data matrix
combined_data_male_trimmed_dm <- as.matrix(combined_data_male_trimmed)

# Create correlation data
correlations_male <- cor(combined_data_male_trimmed_dm, method = "spearman")

# Significance of correlations
p.mat_correlations_male <- cor.mtest(combined_data_male_trimmed_dm)

# Plot

# Open graphical device
pdf("3_OutputFigures/NeutrophilFunctionCorrelations_Male.pdf",
    colormodel = "cmyk")

corrplot(correlations_male, 
         method = "circle", 
         type = "upper",
         order = "original",
         tl.col = "black",
         p.mat = p.mat_correlations_male,
         sig.level = 0.05,
         tl.srt = 45,
         insig = "label_sig",
         pch.col = 'white',
         diag = FALSE,
         col=brewer.pal(n=10, name="RdYlBu"))

# Close Graphical Device
invisible(dev.off())
```

## Females

```{r}
# Repeat same code but for the female data
combined_data_female <- filter(combined_data, Sex == "Female")

# Make data frame with only variables of interest for correlating
combined_data_female_trimmed <- combined_data_female[-c(3:9)]
combined_data_female_trimmed <- data.frame(combined_data_female_trimmed[, -1], row.names = combined_data_female_trimmed$SubjectNumber)

# Convert to data matrix
combined_data_female_trimmed_dm <- as.matrix(combined_data_female_trimmed)

# Create correlation data
correlations_female <- cor(combined_data_female_trimmed_dm, method = "spearman")

# Significance of correlations
p.mat_correlations_female <- cor.mtest(combined_data_female_trimmed_dm)

# Plot

# Open graphical device
pdf("3_OutputFigures/NeutrophilFunctionCorrelations_Female.pdf",
    colormodel = "cmyk")

corrplot(correlations_female, 
         method = "circle", 
         type = "upper",
         order = "original",
         tl.col = "black",
         p.mat = p.mat_correlations_female,
         sig.level = 0.05,
         tl.srt = 45,
         insig = "label_sig",
         pch.col = 'white',
         diag = FALSE,
         col=brewer.pal(n=10, name="RdYlBu"))

# Close Graphical Device
invisible(dev.off())
```

## Individual correlations

OCR Time to Max:
```{r}
# Open graphical device
pdf("3_OutputFigures/Phagocytosis versus OCR Time to Max.pdf",
    colormodel = "cmyk")

## Line graphs
theme_set(theme_bw())

ggplot(combined_data, aes(x= Phagocytosis, y=OCRTimeToMax, color=Sex, shape=Sex)) +
  geom_point(size = 5) + 
  scale_color_manual(values = c("tomato", "steelblue1")) +
  geom_smooth(method=lm, se=FALSE, linetype = "dashed") +
  ylab("OCR Time to Max (minutes)\n") +
  xlab("\nPhagocytosis (MFI)") +
  theme( axis.title.x = element_text(size = 20, face = "bold", color = "black"),
         axis.title.y = element_text(size = 20, face = "bold", vjust = 0.5),
         axis.text.y = element_text(size = 18, color = "black"),
         axis.text.x = element_text(size = 18, color = "black"),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.border = element_rect(colour = "black", fill=NA, size=1),
         legend.text = element_text(size = 16),
         legend.position = c(0.25, 0.85),
         legend.title = element_blank(),
         legend.background = element_rect(fill = "white", 
                                          size = 0.3, 
                                          linetype = "solid",
                                          color = "black"))

# Close Graphical Device
invisible(dev.off())
```

Extract R and p:
```{r}

```


ECAR Time to Max:
```{r}
# Open graphical device
pdf("3_OutputFigures/Phagocytosis versus ECAR Time to Max.pdf",
    colormodel = "cmyk")

## Line graphs
theme_set(theme_bw())

ggplot(combined_data, aes(x= Phagocytosis, y=ECARTimeToMax, color=Sex, shape=Sex)) +
  geom_point(size = 5) + 
  scale_color_manual(values = c("tomato", "steelblue1")) +
  geom_smooth(method=lm, se=FALSE, linetype = "dashed") +
  ylab("ECAR Time to Max (minutes)\n") +
  xlab("\nPhagocytosis (MFI)") +
  theme( axis.title.x = element_text(size = 20, face = "bold", color = "black"),
         axis.title.y = element_text(size = 20, face = "bold", vjust = 0.5),
         axis.text.y = element_text(size = 18, color = "black"),
         axis.text.x = element_text(size = 18, color = "black"),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.border = element_rect(colour = "black", fill=NA, size=1),
         legend.text = element_text(size = 16),
         legend.position = c(0.25, 0.85),
         legend.title = element_blank(),
         legend.background = element_rect(fill = "white", 
                                          size = 0.3, 
                                          linetype = "solid",
                                          color = "black"))

# Close Graphical Device
invisible(dev.off())
```







